---
title: "'Exammarks' Report"
author: "Francesca Vanzo, Leon Schlömmer, Linde Van Saane"
output: pdf_document
date: "February 25th, 2026"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. Introduction

In this report we analyse the data set "2001exammarks.dat", which contains the individual marks each student obtained in questions A1,A2,A3,A4,B1,B2,B3,B4 on the exam course "STAT2001 Probability and Inference". The data set also contains information about the maximum possible grade achievable in each question (which varies based on the question), and the final grade of each student, computed as the sum of the individual marks of each question. The exam is structured so that questions A1 to A4 are easier, and have a maximum total mark of 40 points, while questions B1 to B4 should be more difficult and in total be worth 60 maximum points.

The aim is to identify whether the questions all measure the same skill ( hence how strongly the results in the different questions are related to each other), and whether questions B are actually more difficult than questions A.

## 2. Data analysis

We start by observing the distribution of the grades for each question:

```{r,include=FALSE}
raw <- read.table(
    "~/Desktop/Uni/Adv. analysis/exammarks.dat",
    header    = TRUE, # first line holds column names (A1 A2 ... sum)
    row.names = 1 # first column holds row labels (student IDs / max.)
)
marks <- raw[rownames(raw) != "max.", ]
marks_q <- marks[, colnames(marks) != "sum"]
max_marks <- raw["max.", ]
```

```{r histograms,echo=FALSE,fig.width=10, fig.height=5}
par(mfrow = c(2, 4))
for (q in names(marks_q)) {
 qmax <- as.numeric(max_marks[[q]])
    hist(marks_q[[q]],
        breaks = seq(-0.5, qmax + 0.5, by = 1),
        main = q, xlab = "Marks", col = "lightblue", xaxt = "n"
    )
    axis(1, at = 0:qmax)
}
par(mfrow = c(1, 1))
```

Questions A1,A3,A4 behave how we would expect easier questions to behave, they have all left-skewed distributions, with many students scoring the top marks for each question. When it comes to question B4, which is theoretically a more difficult question,it has most of the students either scoring a high grade, or a very low one, meaning that it was a question for which a student was completely prepared or they wouldn’t be able to score an average mark. On the other hand, questions B2 and B3 present much more variability, with a lower number of students being able to achieve the maximum number of points available. The last two questions are the ones that seem the furthest away from our expectations. The question A2 has a lot more variability in its distribution compared to the other "easy" points of the exam. In fact, most students either scored 6 or 7 points out of the 10 available, and only a low concentration scored more than that. This could indicate that question A2 is not weighted correctly compared to its difficulty, and is much more difficult than the other A questions. Similarly, question B1 is the most left-skewed question out of all the ones we consider, with very few students not scoring the top mark. Hence question B1 also seems to not be labelled correctly based on its difficulty, for reasons inverse to A2.

### -Principal Component Analysis:

```{r,include=FALSE}
pca <- prcomp(marks_q, scale. = TRUE)
summary(pca) # % variance explained
```

```{r,echo=FALSE,fig.width=6, fig.height=3}
plot(pca, type = "l", main = "Screen plot")
```

<br>

```{r pca-loadings-barplots, echo=FALSE, fig.width=8, fig.height=6}
ylim_i <-c(0, min(pca$rotation[, 1]) * 1.1) 
barplot(pca$rotation[, 1],
        main      = paste0("PC1 Loadings"),
        ylab      = "Loading",
        xlab      = "Question",
        col       = "steelblue",
        names.arg = rownames(pca$rotation),
        ylim      = ylim_i
    )
    abline(h = 0, lty = 2, col = "red")
```

Together with the histograms, the PCA loadings give us a good insight on whether the A questions really are easier than the B questions and give weaker students the possibility to reach the pass mark.

If that were the case, here's what we would expect to see:

-   A questions should have lower loadings on PC1, as well as have low variance on the histograms (with most students scoring high)
-   B questions should have higher loadings on PC1 and have more variance

However, the data tells a different story:

Since all questions have loadings with the same sign on PC1, we can assume that PC1 reflects a common skill that is measured across all questions. Let's think of PC1 as a *general ability* dimension.

-   A1 seems to be doing its job. It has a low PC1 loading and low variance with most students scoring high.
-   A2 is a bit of a mixed bag, while it has a relatively low PC1 loading, and not as much variability as other questions, we can see that most students score mediocre. It seems harder than intended.
-   A3 and A4 both have relatively high loadings on PC1, comparable to B questions. They don't seem very accessible.
-   B1 seems to not do its job, we observe that it has the lowest loading among B questions. That, combined with the fact that most students score near the max, B1 is simply too easy to function as a hard question.
-   B2, B3, B4 seem to do their job. High PC1 loadings and they seem to spread the students out the most.

We can furthermore investigate what we have observed with a parallel coordinates plot:

```{r,echo=FALSE,warning=FALSE,message=FALSE,fig.align='right'}
marks_pc <- marks_q
marks_pc$total <- marks$sum
questions <- colnames(marks_q)
breaks <- quantile(marks_pc$total, probs = c(0, 1 / 3, 2 / 3, 1))
marks_pc$performance <- cut(marks_pc$total,
    breaks = breaks,
    labels = c("Low", "Mid", "High"),
    include.lowest = TRUE
)
# Parallel coordinates plot
library(GGally)
ggparcoord(marks_pc,
    columns     = seq_along(questions), # question columns only
    groupColumn = "performance",
    # scale       = "uniminmax", # scale each axis to [0, 1]
    alphaLines  = 0.3
) +
    scale_color_manual(values = c(
        "Low"  = "#d6604d",
        "Mid"  = "#4393c3",
        "High" = "#1a9850"
    )) +
    labs(title="Parallel coordinates plot of the questions:",
        x     = "Question",
        y     = "Scaled mark",
        color = "Performance"
    ) +
    theme_minimal(base_size = 11)+
  theme(plot.title = element_text(hjust = 0.5))
```

This plot gives us another visual representation of the distribution on how the students scored for each question, with marks being divided into 3 categories (low, mid, high). It's clear that questions B2, B3 and B4 behave similarly, with students who performed well in these questions being above average, and students who didn't perform well being under average. The 3 categories for these questions are more distinctly separated vertically compared to the other questions. This would suggest that questions B2-B4 are in fact more difficult and they discriminate performance better than questions A.

In questions A there is more of an overlap between low and mid ranges, and in general they present more variability than B2,B3,B4. A note can be made about A2, which appears to have more variation but also slightly less separation between low, mid and high.

Students with higher grades in each question appear more consistent, especially in section B, while students with a worse performance have a drop in questions B specifically.

Finally, we can also consult a jitter plot of the questions:

```{r jitterplot, fig.width=7, fig.height=4,echo=FALSE}
q_index <- rep(seq_along(questions), each = nrow(marks_q))
max_marks <- raw["max.", ]
mark_vec <- unlist(marks_q)

# Jitter plot — reveals distribution of discrete marks per question
plot(jitter(q_index), jitter(mark_vec),
    xlab = "Questions", ylab = "Mark",
    xaxt = "n", # suppress default x-axis
    pch = 16, cex = 0.7, # solid small dots
    col = adjustcolor("steelblue", alpha.f = 0.5),
    main = "Student marks by question (jittered)"
)

axis(1, at = seq_along(questions), labels = questions)

# Add max-mark reference lines for each question
segments(
    x0 = seq_along(questions) - 0.4,
    x1 = seq_along(questions) + 0.4,
    y0 = as.numeric(max_marks[questions]),
    col = "red", lwd = 2, lty = 2
)

legend("topright", legend = "Max. mark", col = "red", lty = 2, lwd = 2, bty = "n")
```

Here the same conclusions we reached before are once again confirmed. We can see how B1 is clearly the question with the highest concentration of students at the top mark ( indicated in each question by the red line). B2 is the query with the most variability, as we can see its values go from 0 to the top mark and are distributed approximately homogenous.
\newpage

### -Correlation:

Other considerations can be made regarding the relationship between variables. To this aim, we can consult a heat map:

```{r,include=FALSE}
cor_mat <- cor(raw[-1,])
round(cor_mat, 2)
```

```{r,echo=FALSE,fig.width=4, fig.height=4,fig.align='center',message=FALSE,warning=FALSE}
library(corrplot)
corrplot(cor_mat,
    method       = "color", # coloured heatmap
    type         = "upper", # only upper triangle (cleaner)
    tl.cex       = 0.8,
    tl.col = "black" # text label colour
)
```

This plot allows us to observe how each question is correlated to sum. A high correlation between variables means that the variability of one question affects the variability of the "sum" variable ( which indicates the final grade for the exam of each student). Since B questions are supposedly harder and have a larger possible number of points achievable, we expect them to have a high correlation with the sum variable. The A questions are intended for students who don't know as much of the exam material to still be able to pass with the minimum. In this plot we observe how questions B2,B3,B4 all have high correlation with sum (0.89,0.83,0.79 respectively). On the other hand, B1 is not as correlated. Both questions A3 and A4 have a higher correlation with sum than B1, which would mean that they have more "weight" on the final result of the exam compared to it.

## 3. Features of the data

The data set "exammarks" is a data set containing 151 observations for 9 variables (questions A1 to B4 plus the variable "sum" containing the final grade). The data set also has as the first row the values of the maximum achievable mark for each question.

## 4. Conclusions

We have analysed the 9 variables in the data set that indicate the marks that each of the 151 students scored for each question plus the final grade for the exam of the course ""STAT2001 Probability and Inference". The questions were:

-   A1 Basic Probability

-   A2 Two-dimensional densities

-   A3 Distributions derived from the normal distribution

-   A4 Central Limit Theorem and approximation

-   B1 Moment- generating functions

-   B2 Joint, marginal and conditional distribution

-   B3 Multivariate normal distribution

-   B4 Estimation theory

We observed the distribution of each variable and the relationship between each other and between "sum". We wanted to understand whether the questions all measure the same "skill", meaning that we wanted to see how the results of each question were correlated to each other, if it was properly testing the overall knowledge of the student. Furthermore another aim was to test whether questions B were indeed more difficult than questions A.

The results from our work suggest that most questions in the exam measure a common underlying ability related to students’ overall understanding of the course material. This is supported by the PCA results, where all questions have loadings with the same sign on the first principal component, indicating that they are associated with a general ability factor. This means that generally students who perform good in one question will perform well in the others as well. 
However, the analysis also shows that question A2 and B1 do not behave as intended, the problem that arises involves the intended difficulty of the question. While the exam was structured so that questions A are easier and questions B are harder, we have found that in question A2 students scored lower than intended, especially compared to other A questions. Another problem was question B1, where it's clear that too many students scored near the maximum possible grade and the variability was low, meaning that it was too easy compared to the rest of the exam.

Therefore, while generally the questions were able to test an overall knowledge of the students on the material, questions A2 and B1 need to be fixed in order to achieve the desired difference in difficulty between section A and B.

## Appendix

```{r,eval=FALSE}
library(gplots)
library(GGally)
library(ggplot2)
library(corrplot)
# Read the exam marks data
# Row 1 contains the max. possible marks; rows 2+ are student records
raw <- read.table(
    "200120178exammarks.dat",
    header    = TRUE, # first line holds column names (A1 A2 ... sum)
    row.names = 1 # first column holds row labels (student IDs / max.)
)

# Separate the maximum marks row from the student data
max_marks <- raw["max.", ]
marks <- raw[rownames(raw) != "max.", ]
# Drop the sum column — only analyse individual question marks
marks_q <- marks[, colnames(marks) != "sum"]
str(marks_q)
summary(marks_q)
# Quick sanity check
cat("Number of students:", nrow(marks), "\n")
cat("Columns:", paste(colnames(marks), collapse = ", "), "\n")

#Histograms
par(mfrow = c(2, 4))
for (q in names(marks_q)) {
 qmax <- as.numeric(max_marks[[q]])
    hist(marks_q[[q]],
        breaks = seq(-0.5, qmax + 0.5, by = 1),
        main = q, xlab = "Marks", col = "lightblue", xaxt = "n"
    )
    axis(1, at = 0:qmax)
}
par(mfrow = c(1, 1))

markcor <- cor(marks_q)

marks_pc <- marks_q
marks_pc$total <- marks$sum
questions <- colnames(marks_q)
breaks <- quantile(marks_pc$total, probs = c(0, 1 / 3, 2 / 3, 1))
marks_pc$performance <- cut(marks_pc$total,
    breaks = breaks,
    labels = c("Low", "Mid", "High"),
    include.lowest = TRUE
)
# Parallel coordinates plot
ggparcoord(marks_pc,
    columns     = seq_along(questions), # question columns only
    groupColumn = "performance",
    # scale       = "uniminmax", # scale each axis to [0, 1]
    alphaLines  = 0.3
) +
    scale_color_manual(values = c(
        "Low"  = "#d6604d",
        "Mid"  = "#4393c3",
        "High" = "#1a9850"
    )) +
    labs(title="Parallel coordinates plot of the questions:",
        x     = "Question",
        y     = "Scaled mark",
        color = "Performance"
    ) +
    theme_minimal(base_size = 11)+
  theme(plot.title = element_text(hjust = 0.5))

q_index <- rep(seq_along(questions), each = nrow(marks_q))
max_marks <- raw["max.", ]
mark_vec <- unlist(marks_q)

# Jitter plot — reveals distribution of discrete marks per question
plot(jitter(q_index), jitter(mark_vec),
    xlab = "Questions", ylab = "Mark",
    xaxt = "n", # suppress default x-axis
    pch = 16, cex = 0.7, # solid small dots
    col = adjustcolor("steelblue", alpha.f = 0.5),
    main = "Student marks by question (jittered)"
)

axis(1, at = seq_along(questions), labels = questions)

# Add max-mark reference lines for each question
segments(
    x0 = seq_along(questions) - 0.4,
    x1 = seq_along(questions) + 0.4,
    y0 = as.numeric(max_marks[questions]),
    col = "red", lwd = 2, lty = 2
)
legend("topright", legend = "Max. mark", col = "red", lty = 2, lwd = 2, bty = "n")

cor_mat <- cor(raw[-1,])
round(cor_mat, 2)
corrplot(cor_mat,
    method       = "color", # coloured heatmap
    type         = "upper", # only upper triangle (cleaner)
    tl.cex       = 0.8,
    tl.col = "black" # text label colour
)

#PCA:
pca <- prcomp(marks_q, scale. = TRUE)
summary(pca) # % variance explained
pca$rotation # loadings
plot(pca, type = "l", main = "Scree plot")
biplot(pca, scale = 0)

#Comparison between A and B:
A_mean <- apply(marks_q[, 1:4], 2, mean)
B_mean <- apply(marks_q[, 5:8], 2, mean)
A_mean
B_mean
t.test(rowSums(marks_q[, 1:4]), rowSums(marks_q[, 5:8]))
marks_q$A_total <- rowSums(marks_q[, 1:4])
marks_q$B_total <- rowSums(marks_q[, 5:8])

plot(marks_q$A_total, marks_q$B_total,
    xlab = "Total A marks", ylab = "Total B marks",
    pch = 19, col = "darkblue",
    main = "A vs B marks per student"
)
abline(lm(B_total ~ A_total, data = marks_q), col = "red")
# Clean up: remove temporary columns so marks_q stays clean for later use
marks_q$A_total <- NULL
marks_q$B_total <- NULL

# Fit a linear model: sum ~ A1 + A2 + A3 + A4 + B1 + B2 + B3 + B4
model <- lm(sum ~ A1 + A2 + A3 + A4 + B1 + B2 + B3 + B4, data = marks)
summary(model)
```
